Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2022-06-29 manuscript.tex
Improvements

\begin{itemize}
\tightlist
\item
  A comparison of and guidance for the use of two software packages,\\
\item
  Guidance on prior specification,
\item
  Practical instructions for Bayes factor model comparison.
\end{itemize}

@@

\caption{Panel A shows the response times in seconds for two conditions of the number classification task (i.e., digits 3 and 4). The grey lines represent the individual effects, and the blue line the aggregated effect. Panel B shows how individual estimates are shrunken towards the general trend. Sample represents the observed effect per participant in seconds, while Model represents the estimated effects per participant. }\label{fig:figure1}
\end{figure}

@@

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend (Rouder, Morey, \& Pratte, 2013; Haaf \& Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, 2021). This is shown in Panel A of Figure \ref{fig:figure1} where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a symbolic distance effect experiment repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. Therefore, the observations are not independent as we assume if we perform analysis on aggregated data. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder \& Lu, 2005).

@@

\caption{Hierarchical structure of the normal and lognormal models for the number classification task.}\label{fig:hierarchyplot}
\end{figure}

@@

The simulation proceeds as follows: First, one random draw from each prior distribution is taken. This is plugged into the individual effect distributions (Equation (2)). Then, the likelihood of the model, such as specified in Equations (1) and (3), is used to simulate observations per condition. This means that for \emph{n} participants, \emph{n} random values for the model parameters are generated based on the priors, which will result in the \emph{k} observations per \emph{j} conditions. This process is repeated \emph{m} times, where \emph{m} is the number of simulations drawn. Note that the more simulations are conducted, the closer one gets to the full picture because enough samples from the prior distribution are needed. Finally, we can visualize the simulated data. Figure \ref{fig:priorpredfigurepanel} shows prior predictions for the key effects from the number classification task. For this task, we mainly care about the side effect and digit effects corresponding to the \(\beta\)s and the \(\delta\)s from the model. Here, the prior predictions for the average side effect and the average digit effect across participants are depicted.

@@

\normalsize If you do not specify a prior for a certain parameter, the default prior as explained in the prior section will be applied. It is possible to save all priors in an R \texttt{data.frame} and provide these in the \texttt{brm} function, this and the full code with explanation for fitting the full symbolic distance effect model using \texttt{brms} is shown in our \href{https://www.dropbox.com/s/mthphl3ce2kk0xx/Appendix-D--Fit-BHM-in-brms.pdf?dl=0}{Online Supplement D: \emph{Bayesian Hierarchical Modeling in brms}}.\footnote{In our example of the Number Classification Task we are applying a truncated normal prior distribution for only one parameter. This requires a slightly different \texttt{brms} formula and prior specification, explained in the Online Supplement.}

\hypertarget{estimation-of-the-number-classification-task}{%
\section{Estimation of the Number Classification Task}\label{estimation-of-the-number-classification-task}}

@@

For our example, we will use the data set by Rouder et al.~(2005). The data are available under a free license on GitHub \url{https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5}. The data set contains observations from 54 participants who took part in a number classification task. Each participant performed 360 trials in six blocks. For the analysis, only the correct responses are included. To account for practice effects, the first 25 trials of the first block are removed from the analysis. The trials after the breaks between the blocks are also removed from the analysis to correct for lack of concentration. In addition, trials with very fast RTs (\textless{} 250 MS) and very slow RTs (\textgreater{} 2000 MS) are excluded. Two participants were excluded because they did not respond properly to the task. Similar criteria were applied by Haaf, Klaassen, and Rouder (2019). This results in 52 participants and a total of 17031 observations. In the following two sections we will describe the model fitting and parameter estimation procedure for the normal and log-normal model, respectively.

@@

\begin{itemize}
\tightlist
\item
  Document layout: \emph{rmarkdown} {[}Version 1.16; Xie, Allaire, and Grolemund (2018){]}, \emph{papaja} {[}Version 0.1.0.9842; Aust and Barth (2018){]}, \emph{knitr} {[}Version 1.25; Xie (2015){]}, \emph{kableExtra} {[}Version 1.1.0; Zhu (2019){]}
\item
  Data analysis and cleaning: \emph{LaplacesDemon} {[}Version 16.1.4; Statisticat and LLC. (2020){]}, \emph{plyr} {[}Version 1.8.4; Wickham, François, Henry, and Müller (2019); Wickham (2011){]}, \emph{dplyr} {[}Version 0.8.3; Wickham, François, Henry, and Müller (2019){]}, \emph{readr} {[}Version 1.3.1; Wickham, Hester, and Francois (2018){]}, \emph{crch} {[}Version 1.0.4; Messner, Mayr, and Zeileis (2016); Messner, Mayr, and Zeileis (2017); Messner, Zeileis, Broecker, and Mayr (2014); Messner, Mayr, Zeileis, and Wilks (2014){]}, \emph{foreign} {[}Version 0.8.71; R Core Team (2018){]}, \emph{Matrix} {[}Version 1.2.17; Bates and Maechler (2019){]}, \emph{tidyr} {[}Version 1.0.0; Wickham and Henry (2019){]}
\item
  Model fit and comparison: \emph{rstan} {[}Version 2.19.2; Stan Development Team (2019a){]}, \emph{brms} {[}Version 2.13.0; Bürkner (2017); Bürkner (2018){]}, \emph{BayesFactor} {[}Version 0.9.12.4.2; Morey and Rouder (2018){]}, \emph{bridgesampling} {[}Version 1.0.0; Gronau, Singmann, and Wagenmakers (2020){]}
\item
  Visualization: \emph{coda} {[}Version 0.19.3; Plummer, Best, Cowles, and Vines (2006){]}, \emph{lattice} {[}Version 0.20.38; Sarkar (2008){]}, \emph{ggplot2} {[}Version 3.3.2; Wickham (2016){]}, \emph{Rmisc} {[}Version 1.5; Hope (2013){]}, \emph{devtools} {[}Version 2.2.1; Wickham, Hester, and Chang (2019){]}, \emph{gghalves} {[}Version 0.1.0; Tiedemann (2020){]}, \emph{bayesplot} {[}Version 1.7.1; Gabry, Simpson, Vehtari, Betancourt, and Gelman (2019){]}, \emph{gridExtra} {[}Version 2.3; Auguie (2017){]}, \emph{tibble} {[}Version 3.0.2; Müller and Wickham (2019){]}, \emph{ggmcmc} {[}Version 1.4.1; Fernández-i-Marín (2016){]}, \emph{cowplot} {[}Version 1.0.0; Wilke (2019){]}, \emph{ggridges} {[}Version 0.5.1; Wilke (2018){]}, \emph{circlize} {[}Version 0.4.9; Gu, Gu, Eils, Schlesner, and Brors (2014){]}, \emph{StanHeaders} {[}Version 2.19.0; Stan Development Team (2019b){]}, \emph{unikn} {[}Version 0.2.0; Neth and Gradwohl (2019){]}
\item
  Other: \emph{Rcpp} {[}Version 1.0.3; Eddelbuettel and François (2011); Eddelbuettel and Balamuta (2017){]}, \emph{usethis} {[}Version 1.5.1; Wickham and Bryan (2019){]}
\end{itemize}

@@

@@

>>>>

\begin{itemize}
\tightlist
\item
  A comparison of and guidance for the use of two software packages,
\item
  Guidance on prior specification,
\item
  Practical instructions for Bayes factor model comparison.
\end{itemize}

@@

\caption{Panel A shows the response times in seconds for two conditions of the digit classification task (i.e., digits 3 and 4). The grey lines represent the individual effects, and the blue line the aggregated effect. Panel B shows how individual estimates are shrunken towards the general trend. Sample represents the observed effect per participant in seconds, while Model represents the estimated effects per participant. }\label{fig:figure1}
\end{figure}

@@

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend (Rouder, Morey, \& Pratte, 2013; Haaf \& Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, 2021). This is shown in Panel A of Figure \ref{fig:figure1} where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. Therefore, the observations are not independent as we assume if we perform analysis on aggregated data. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder \& Lu, 2005).

@@

\caption{Hierarchical structure of the normal and lognormal models for the digit classification task.}\label{fig:hierarchyplot}
\end{figure}

@@

The simulation proceeds as follows: First, one random draw from each prior distribution is taken. This is plugged into the individual effect distributions (Equation (2)). Then, the likelihood of the model, such as specified in Equations (1) and (3), is used to simulate observations per condition. This means that for \emph{n} participants, \emph{n} random values for the model parameters are generated based on the priors, which will result in the \emph{k} observations per \emph{j} conditions. This process is repeated \emph{m} times, where \emph{m} is the number of simulations drawn. Note that the more simulations are conducted, the closer one gets to the full picture because enough samples from the prior distribution are needed. Finally, we can visualize the simulated data. Figure \ref{fig:priorpredfigurepanel} shows prior predictions for the key effects from the digit classification task. For this task, we mainly care about the side effect and digit effects corresponding to the \(\beta\)s and the \(\delta\)s from the model. Here, the prior predictions for the average side effect and the average digit effect across participants are depicted.

@@

\normalsize If you do not specify a prior for a certain parameter, the default prior as explained in the prior section will be applied. It is possible to save all priors in an R \texttt{data.frame} and provide these in the \texttt{brm} function, this and the full code with explanation for fitting the full symbolic distance effect model using \texttt{brms} is shown in our \href{https://www.dropbox.com/s/mthphl3ce2kk0xx/Appendix-D--Fit-BHM-in-brms.pdf?dl=0}{Online Supplement D: \emph{Bayesian Hierarchical Modeling in brms}}.\footnote{In our example of the digit classification task we are applying a truncated normal prior distribution for only one parameter. This requires a slightly different \texttt{brms} formula and prior specification, explained in the Online Supplement.}

\hypertarget{estimation-of-the-digit-classification-task}{%
\section{Estimation of the Digit Classification Task}\label{estimation-of-the-digit-classification-task}}

@@

For our example, we will use the data set by Rouder et al.~(2005). The data are available under a free license on GitHub \url{https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5}. The data set contains observations from 54 participants who took part in a digit classification task. Each participant performed 360 trials in six blocks. For the analysis, only the correct responses are included. To account for practice effects, the first 25 trials of the first block are removed from the analysis. The trials after the breaks between the blocks are also removed from the analysis to correct for lack of concentration. In addition, trials with very fast RTs (\textless{} 250 MS) and very slow RTs (\textgreater{} 2000 MS) are excluded. Two participants were excluded because they did not respond properly to the task. Similar criteria were applied by Haaf, Klaassen, and Rouder (2019). This results in 52 participants and a total of 17031 observations. In the following two sections we will describe the model fitting and parameter estimation procedure for the normal and log-normal model, respectively.

@@

\begin{itemize}
\tightlist
\item
  Document layout: \emph{rmarkdown} {[}Version 1.16; Xie, Allaire, and Grolemund (2018){]}, \emph{papaja} {[}Version 0.1.0.9842; Aust and Barth (2018){]}, \emph{knitr} {[}Version 1.25; Xie (2015){]}, \emph{kableExtra} {[}Version 1.1.0; Zhu (2019){]}
\item
  Data analysis and cleaning: \emph{LaplacesDemon} {[}Version 16.1.4; Statisticat and LLC. (2020){]}, \emph{plyr} {[}Version 1.8.4; Wickham, François, Henry, and Müller (2019); Wickham (2011){]}, \emph{dplyr} {[}Version 0.8.3; Wickham, François, Henry, and Müller (2019){]}, \emph{readr} {[}Version 1.3.1; Wickham, Hester, and Francois (2018){]}, \emph{crch} {[}Version 1.0.4; Messner, Mayr, and Zeileis (2016); Messner, Mayr, and Zeileis (2017); Messner, Zeileis, Broecker, and Mayr (2014); Messner, Mayr, Zeileis, and Wilks (2014){]}, \emph{foreign} {[}Version 0.8.71; R Core Team (2018){]}, \emph{Matrix} {[}Version 1.2.17; Bates and Maechler (2019){]}, \emph{tidyr} {[}Version 1.0.0; Wickham and Henry (2019){]}, \emph{truncnorm} {[}Version 1.0.8; Mersmann, Trautmann, Steuer, and Bornkamp (2018){]}
\item
  Model fit and comparison: \emph{rstan} {[}Version 2.19.2; Stan Development Team (2019a){]}, \emph{brms} {[}Version 2.13.0; Bürkner (2017); Bürkner (2018){]}, \emph{BayesFactor} {[}Version 0.9.12.4.2; Morey and Rouder (2018){]}, \emph{bridgesampling} {[}Version 1.0.0; Gronau, Singmann, and Wagenmakers (2020){]}
\item
  Visualization: \emph{coda} {[}Version 0.19.3; Plummer, Best, Cowles, and Vines (2006){]}, \emph{lattice} {[}Version 0.20.38; Sarkar (2008){]}, \emph{ggplot2} {[}Version 3.3.2; Wickham (2016){]}, \emph{Rmisc} {[}Version 1.5; Hope (2013){]}, \emph{devtools} {[}Version 2.2.1; Wickham, Hester, and Chang (2019){]}, \emph{gghalves} {[}Version 0.1.0; Tiedemann (2020){]}, \emph{bayesplot} {[}Version 1.7.1; Gabry, Simpson, Vehtari, Betancourt, and Gelman (2019){]}, \emph{gridExtra} {[}Version 2.3; Auguie (2017){]}, \emph{tibble} {[}Version 3.0.2; Müller and Wickham (2019){]}, \emph{ggmcmc} {[}Version 1.4.1; Fernández-i-Marín (2016){]}, \emph{cowplot} {[}Version 1.0.0; Wilke (2019){]}, \emph{ggridges} {[}Version 0.5.1; Wilke (2018){]}, \emph{circlize} {[}Version 0.4.9; Gu, Gu, Eils, Schlesner, and Brors (2014){]}, \emph{StanHeaders} {[}Version 2.19.0; Stan Development Team (2019b){]}, \emph{unikn} {[}Version 0.2.0; Neth and Gradwohl (2019){]}, \emph{ggbeeswarm} {[}Version 0.6.0; Clarke and Sherrill-Mix (2017){]}
\item
  Other: \emph{Rcpp} {[}Version 1.0.3; Eddelbuettel and François (2011); Eddelbuettel and Balamuta (2017){]}, \emph{usethis} {[}Version 1.5.1; Wickham and Bryan (2019){]}
\end{itemize}

@@

\leavevmode\hypertarget{ref-R-ggbeeswarm}{}%
Clarke, E., \& Sherrill-Mix, S. (2017). \emph{Ggbeeswarm: Categorical scatter (violin point) plots}. Retrieved from \url{https://CRAN.R-project.org/package=ggbeeswarm}

@@

\leavevmode\hypertarget{ref-R-truncnorm}{}%
Mersmann, O., Trautmann, H., Steuer, D., \& Bornkamp, B. (2018). \emph{Truncnorm: Truncated normal distribution}. Retrieved from \url{https://CRAN.R-project.org/package=truncnorm}
